---
title: "Mapping Internet Access in Multnomah County"
author: "Charles Hanks"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
census_key = "a8150f07baa78cbba0671478efbfa37d0d2f9aa4"
```

# Introduction 

Suppose my client is the Multnomah County Library (MCL). Following the approval of Measure 26-211 by voters in 2020, the county is funding renovation of 8 of the 19 libraries and creating a new flagship library at 7905 SE Holgate to serve East Portland. 

As our daily lives have become increasingly digital, access to adequate information technology has become a necessity. The COVID-19 pandemic accelerated this shift, where video conferencing became the norm for so many of us to go to school, the doctor, and work. 

However, not everyone has access to the internet at home in 2023.  Our libraries provide access to information technology to everyone at no cost. The Multnomah County Library has asked that I provide a report addressing the following questions:  

* Where in the county are there the most people without broadband internet access at home? 
* Are the libraries' locations convenient for those who need access to information technology?  

I used the 2021 American Community Survey to answer who how many people not have have access to the internet at home, where they live, and created visualization to present my findings to the Library Advisory Board. 

# My Findings 

* Over 20,000 people in Multnomah County do not have access to broadband internet at home. 
* The largest concentrations of people without internet at home are in Downtown (15.6%) and Old Town Portland (17.5%), followed by the Piedmont Neighborhood (8%). 
* With the Central and Holgate Libraries closed for construction through 2023, there is currently a gap in service in providing Downtown and East Portland with access to information technology. 
* MCL plans to open a pop-up technology resource center at 510 SE Washington, however no opening date has been announced. 


# Deliverables 

* [Static Map of Multnomah County](https://chanks06.github.io/multco_internet_map.pdf)
* [Interactive Map of Multnomah County](https://chanks06.github.io/multco_leaflet.html)  
  
\
\
\
\

# Code 

Loading packages:  
```{r}
library(tidyverse)
library(maps)
library(sf)
library(tigris)
library(tidycensus)
library(tmap)
library(leaflet)
library(htmltools)
library(htmlwidgets)
```

Authenticating US Census API key: 
```{r, include = T}
options(tigris_use_cache = TRUE)
#getting access to census day with my API key 
census_api_key(census_key, install = TRUE, overwrite = TRUE)
```


Loading the American Community Survey data from the Census bureau: 
```{r}
#grabbing variables from 2021 ACS 1-year estimates (data collected Jan 1 2021 - Dec 31 2021)
acs_vars_2021 = load_variables(year = 2021, dataset = "acs1")

#searching for variables with internet 
acs_vars_2021 %>% filter(grepl("internet", concept, ignore.case = T)) #| grepl("computer", concept, ignore.case = T))

#looks like I will need to take into account the population per tract in order to calculate percent of those without internet at home. 
acs_vars_2021 %>% filter(grepl("population", concept, ignore.case = T))
```

Potenial variables to use: 

B01003_001 total population 
B28002_013 no internet access 
B28002_004 with broadband of any type 
B28003_006 no computer
B28003_004 has computer with broadband internet 
B01003_001 total population 

```{r}
this.year = 2021

#downloading census tract shapefile of multnomah county 
or_tracts = tracts(state = 'OR', county = 'Multnomah', cb = TRUE, year = this.year)

pop_and_no_internet = get_acs(geography = "tract", year=this.year,
              state = "OR", county = "Multnomah",
              variables = c(total_pop = "B01003_001",
                            no_internet = "B28002_013")) %>% 
                    mutate(AFFGEOID=paste0("1400000US", GEOID)) #creating AFFGEOID id from GEOID in order to join data to tract spatial data

pop_no_int = as.data.frame(pop_and_no_internet)[,c(1,3:4)] %>% 
    pivot_wider(names_from = variable, values_from = estimate) %>% #making the data frame tidy 
      mutate(perc_no_int = round((no_internet/total_pop)*100,2), #creating variable that is percentage of population in tract that does not have internet
              AFFGEOID=paste0("1400000US", GEOID))

#summary data 
pop_no_int %>% drop_na() %>% summarize(sum_pop = sum(total_pop), sum_no_internet = sum(no_internet))

pop_no_int %>% arrange(desc(perc_no_int))
# GEOIDs 41051005103 has estimated 17 % no internet at home, 41051010601 has almost 16% 


#joining this ACS data with tract spatial data
pop_no_int_tracts = geo_join(or_tracts, pop_no_int, by_sp = "AFFGEOID", by_df = "AFFGEOID")

#first visualization using
map2 = tm_shape(pop_no_int_tracts) + 
  tm_fill("perc_no_int", style = "cont", palette = "Oranges", title = "") + 
  tm_style("grey") + 
  tm_legend(bg.color = "white", bg.alpha = 0.6, position = c("left", "bottom")) + 
  tm_layout(title = "Percent of Households in Multnomah\nCounty without Broadband Internet",
            title.position = c("center", "top"),
            title.size = 1,
            fontfamily = "sans", fontface = "bold") +   
  tm_credits("Top 5 Neighborhoods    ",size = .8, position = c("RIGHT", "BOTTOM")) + 
  tm_credits("Old Town: 17.46 %\nSW Downtown: 15.69%\nSE Downtown: 8.19%\nPiedmont: 8.14%\nHazelwood 6.91%", size = .7, position = c("RIGHT", "BOTTOM")) + 
  tm_credits("Data source: American Community Survery 2021", size = .6, fontface = "italic", position = c("LEFT", "BOTTOM")) 

  
pop_no_int_tracts = pop_no_int_tracts %>% 
                      arrange(desc(perc_no_int)) %>% 
                          mutate(name = case_when(TRACTCE == "005103" ~ "Old Town",
                                                  TRACTCE == "010601" ~ "SW Downtown", 
                                                  TRACTCE == "010602" ~ "SE Downtown", 
                                                  TRACTCE == "003701" ~ "Piedmont", 
                                                  TRACTCE == "008203" ~ "Hazelwood"))


```
Creating an interactive graphic using leaflet: 
```{r}
pal<-colorNumeric("Reds", domain=0:ceiling(max(pop_no_int_tracts$perc_no_int, na.rm=TRUE)))

popup<-paste("Tract: ", as.character(substring(pop_no_int_tracts$GEOID.x, 6, 11)), "<br>",
             "Percentage of Households without broadband internet: ", as.character(pop_no_int_tracts$perc_no_int))

library(leafletCN)


multco_leaflet = leaflet()%>%
  addProviderTiles("CartoDB.Positron")%>%
  addPolygons(data=pop_no_int_tracts,
              fillColor= ~pal(pop_no_int_tracts$perc_no_int),
              fillOpacity = .5,
              weight =.5,
              smoothFactor = 0.2,
              popup = popup) %>% 
  addMarkers(lat = 45.519702018355936, lng = -122.6831405693095, popup = "Multnomah County Central Library - 801 SE 10th Ave (closed for construction)") %>% 
  addMarkers(lat = 45.582959138612935, lng = -122.68644127317535, popup = "Multnomah County Kenton Library - 8226 NE Denver Ave") %>% 
  addMarkers(lat = 45.490583320668875, lng =  -122.582268235629, popup = "Multnomah Count Holgate LIbrary - 7905 SE Holgate Blvd (closed for construction)") %>%
  addControl(
    position = "bottomright", 
    html = "<h4> Percent of Households without Access to the Internet in Multnomah County</h4>"
  )

#exporting interactive map as html: 
#saveWidget(multco_leaflet, file ="multco_leaflet.html")
```

