---
title: "Scraping & Wrangling in R:\nMultnomah County Library Hours"
author: "Charles Hanks" 
output: html_document
date: "2023-02-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# When is the Library Open? 

I go to the public library often to print or work in a quiet space without feeling the the obligation of commerce. This is a simple demonstration of scraping useful information from a public website, and then organizing it into a table in R. 

"Having fun isn't hard, when you've got a library card!" 

-Arthur and Friends 

```{r}
library(tidyverse)
library(rvest)
library(stringr)
library(lubridate)
library(hms)
```

Scraping data from Multnomah County website:
```{r}
#url to scrape from
result = read_html('https://multcolib.org/hours-and-locations')

#library names and addresses 
lib_names = result %>% 
  html_nodes('.p-name') %>% html_text()

lib_addresses = result %>% 
  html_nodes('.p-street-address') %>% html_text()

```

Creating the table: 
```{r}
#hours
hours = result %>% 
  html_nodes('.hours') %>% 
    html_text()

#removing html line break characters 
hours = (str_split(hours, "\r\n")) 

#removing "Weekly Hours" string"
hours = lapply(hours, function(x) x[-1])

hours = data.frame(hours)

#temporary colnames
colnames(hours) = c(1:22)

#removing day of week from time interval 
hours = hours %>% 
  mutate(across(everything(),~str_remove(.,"\\w+day")))


hours2 = hours %>% gather(key = "key", value = "value", 1:22)

#adding day of week per open hours 
hours2$day = rep(c("monday", "tuesday", "wednesday", "thursday", "friday", "saturday", "sunday", NA),22)

#spreading out days of week
hours3 = hours2 %>% 
            spread(key = day, value = value) 

#converting index to double 
hours3$key = as.numeric(hours3$key)

hours3 = hours3[,c("key","monday","tuesday","wednesday","thursday","friday","saturday","sunday")] %>% arrange(key)

#bringing in library names
hours3$library = lib_names

#renaming key as id 
hours3 = hours3 %>% rename(id = key)

ds = hours3

#adding library addresses: 
ds$address = lib_addresses

ds = ds[,c("id", "library","address","monday","tuesday","wednesday","thursday","friday","saturday","sunday")]
        
ds

```


